%%% QUESTION %%%

\begin{problem}{Problem 2: Linear programming problem}
In a plant three products, R, S, and T are made in two process stages A and B.
To make a product the following time in each process stage is required:
\begin{itemize}
  \item 1 tonne of R: 3 hours in stage A plus 2 hours in stage B.
  \item 1 tonne of S: 2 hours in stage A and 2 hours in stage B.
  \item 1 tonne of T: 1 hour in stage A and 3 hours in stage B.
\end{itemize}
During one year, stage A has 7200 hours and stage B has 6000 hours available production time.
The rest of the time is needed for maintenance. It is required that the available production time should be fully utilized in both stages.

\medskip The profit from the sale of the product is:
\begin{itemize}
  \item R: 100 NOK per tonne.
  \item S: 75 NOK per tonne.
  \item T: 55 NOK per tonne.
\end{itemize}
We wish to mazimize the yearly profit.


\medskip (a) Formulate this as an LP problem on standard form.

\medskip (b) Which basic feasible points exist?

\medskip (c) Find the solution by checking the KKT conditions at all the feasible points found in (b).

\medskip (d) Formulate the dual problem for the LP in (a).

\medskip (e) Show that the optimal objective function value for the LP in (a) equals the optimal objective
function value for the dual problem in (d) by showing that $c^{\top}x^* = b^{\top}\lambda^*$.

\medskip (f) If you can make either stage A or stage B more available, which should you choose?


\end{problem}

%%% SOLUTION %%%

\SUBTASK{a}{Standard form}

\begin{align*}
  \min \qquad         & -c^{\top}x \qquad \text{s.t.} \qquad Ax = b,\ x \geq 0 \\
  \text{where} \qquad & c = \M{100                                             \\ 75 \\ 65},\  x = \M{x_1 \\ x_2 \\ x_3},\  A = \M{3 & 2 & 1 \\ 2 & 2 & 3},\ b = \M{7200  \\ 6000}
\end{align*}

\SUBTASK{b}{Feasible points}

The feasible points are the intersections between the restrictions. We can find these by picking two indices $i, j \in \{1, 2, 3\}$ and solving $\M{A_i & A_j}\M{x_i  \\ x_j } = b$

\medskip
\begin{minipage}[c]{0.55\textwidth}
  With $i, j = 1, 2$:
  \[
    \MGJ{cc|c}{3 & 2 & 7200 \\ 2 &  2 & 6000} \rightarrow \MGJ{cc|c}{1 & 2 /3 & 2400 \\ 0 & 2 /3 & 1200} \rightarrow \MGJ{cc|c}{1 & 0 & 1200 \\ 0 & 1 & 1800}
    .\]
\end{minipage}
\begin{minipage}[c]{0.45\textwidth}
  \[
    x^{(1)} = \M{1200 \\ 1800 \\ 0}
    .\]
\end{minipage}

\bigskip
\begin{minipage}[c]{0.55\textwidth}
  With $i, j = 1, 3$:
  \[
    \MGJ{cc|c}{3 & 1 & 7200 \\ 2 & 3 & 6000} \rightarrow \MGJ{cc|c}{1 & 1 / 3 & 2400 \\ 0 & 7 / 3 & 1200} \rightarrow \MGJ{cc|c}{1 & 0 & 15600 / 7 \\ 0 & 1 & 3600 / 7}
    .\]
\end{minipage}
\begin{minipage}[c]{0.45\textwidth}
  \[
    x^{(2)} = \M{2400 \\ 0 \\ 1200}
    .\]
\end{minipage}

\bigskip
\begin{minipage}[c]{0.55\textwidth}
  With $i, j = 2, 3$:
  \[
    \MGJ{cc|c}{2 & 1 & 7200 \\ 2 & 3 & 6000} \rightarrow \MGJ{cc|c}{1 & 1 /2  & 3600 \\ 0 & 2 & -1200} \rightarrow \MGJ{cc|c}{1 & 0 & 3900 \\ 0 & 1 & -600}
    .\]
\end{minipage}
\begin{minipage}[c]{0.45\textwidth}
  \[
    x^{(3)} = \M{0 \\ 3900 \\ -600}
    .\]
\end{minipage}

\SUBTASK{c}{Check KKT}

Look at $x^{(1)}$:

\medskip Find s such that $s^{\top}x^{(1)} = 0$:
\[
  \M{s_1 & s_2 & s_3}\M{1200 \\ 1800 \\ 0}= 0 \implies s = \M{0 \\ 0 \\ s_3}
  .\]
Now check KKT-1: $A^{\top}\lambda^*+ s^*=c$:
\[
  \M{3 & 2 \\ 2 & 2 \\ 1 & 3}\M{\lambda_1 \\ \lambda_2} + \M{0 \\ 0 \\ s_3} = -\M{100 \\ 75 \\ 55} \implies \MGJ{ccc|c}{3 & 2 & 0 & -100 \\ 2 & 2 & 0 & -75 \\ 1 & 3 & 1 & -55} \rightarrow \MGJ{ccc|c}{1 & \frac{2}{3} & 0 & \frac{-100}{3} \\ 0 & \frac{2}{3} & 0 & \frac{-25}{3} \\ 0 & \frac{7}{3} & 1 & \frac{-65}{3}}
  \rightarrow \M{\lambda_1 \\ \lambda_2 \\ s_3} = \M{-25 \\ -25 /2  \\ 15 /2}
.\] 

This is a valid solution. Since satisfying the KKT conditions is both necessary and sufficient for LP problems, we do not need to check the other points. Anyway, 

\medskip Look at $x^{(2)}$:

\medskip Find s such that $s^{\top}x^{(2)}$:
\[
  \M{s_1 & s_2 & s_3}\M{2400 \\ 0 \\ 1200} = 0 \implies s = \M{0 \\ s_2 \\ 0}
.\] 
Now check KKT-1: $A^{\top}\lambda^*+s^* = c$: 
\[
  \M{3 & 1 \\ 2 & 2 \\ 1 & 3}\M{\lambda_1 \\ \lambda_2}+ \M{0 \\ s_2 \\ 0} = - \M{100 \\ 75 \\ 55} \implies \MGJ{ccc|c}{3 & 2 & 0 & -100 \\ 2 & 2 & 1 & -75 \\ 1 & 3 & 0 & -55} 
  \rightarrow \MGJ{ccc|c}{1 & \frac{2}{3} & 0 & -100 / 3 \\ 0 & \frac{2}{3} & 1 & -25 /3 \\ 0 & \frac{7}{3} & 0 & -65 /3 }
  \rightarrow m\MGJ{ccc|c}{1 & 0 & -1 & -25 \\ 0 & 1 & \frac{3}{2} & -25 /2  \\ 0 & 0 & \frac{-7}{2} & 15 /2}
.\] 
\[
  \rightarrow \MGJ{ccc|c}{1 & 0 & 0 & -190 / 7 \\ 0 & 1 & 0 & -65 / 7 \\ 0 & 0 & 1 & -15 /7}
  \implies \M{\lambda_1 \\ \lambda_2 \\ s_2} = \M{-190 / 7\\ -65 /7 \\ -15 / 7}
.\] 

We see that $x^{(2)}$ does not satisfy the KKT conditions because $s_2 < 0$. 

\medskip No need to look at $x^{(3)}$, because it is not greater than 0, breaking a constraint. 

The optimal solution is now $f(x^{(1)}) = -\M{100 & 75 & 55}^{\top}\M{1200 & 1800 & 0} = -255000$

\SUBTASK{d}{Dual}

The dual is $\max b^{\top}\lambda\qquad\text{s.t.}\qquad A^{\top}\lambda \leq c$

\SUBTASK{e}{Primal and dual are equal}

\begin{minipage}[t]{0.3\textwidth}
  \centering
  Using: 
  \begin{align*}
    A^{\top}\lambda^* + s^* &= c \\
    Ax^* &= b \\ 
    s^{*T}x^* &= 0
  \end{align*}
\end{minipage}%
\hfill
\begin{minipage}[t]{0.7\textwidth}
  \centering Gives:
  \begin{align*}
    c^{\top}x^* &= (A^{\top}\lambda^*+s^*)^{\top}x^*  \\ 
    &= \lambda^{*\top}Ax^* + s^{*\top}x^* \\ 
    &= \lambda^{*\top}b + s^{*\top}x^*  \\ 
    &= \lambda^{*\top}b + 0 \\ 
    &= b^{\top}\lambda^*
  \end{align*}
\end{minipage}

\SUBTASK{d}{Increase availability}

To decide which stage to increase availability, we only need to look at the shadow price $(\lambda^*)$. The value of these show how much we 
gain by increasing the availability. Since 25 = $|\lambda_1^*|>|\lambda_2^*|$ = 25/2, we choose to relax constraint 1 (increase availability of A). One unit relaxation (one hour added) improves the
objective by $\lambda_1 = 25$. So the new optimal value is -255025. 
